\pkg{cubble}: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2022-06-08 1-intro.Rmd
proofread

Spatio-temporal data has a spatial component referring to the location of each observation and a temporal component that is recorded at regular or irregular time intervals. It may also include multiple variables measured at each spatial and temporal values. With spatio-temporal data, one can fix the time to explore the spatial features of the data, fix the spatial location/s to explore temporal aspects, or dynamically explore the space and time simultaneously. 
<!-- Underlying all these tasks, there is a data representation that enables the spatio-temporal data to be sliced and diced differently.  -->
In order to computationally explore the spatial, temporal and spatio-temporal faces of such data, the data needs to be stored and represented under a specific data object that allows the user to query, group and dissect all the data faces.

@@

<!-- **I would aligned the examples of spatio-temporal data with those discuss in the st books. See Paula Moraga for example** -->

<!-- For spatio-temporal data analysis, merging all these sources can be challenging and joining these tables can go wrong because of the slightest unmatched of the linking variable from different sources.  -->

@@

<!-- there is a strong need for such tools given the ubiquitous spatial and temporal data collected in the society, i.e., climate observations from weather stations, air quality variables from air monitoring stations. The demand for such tools can also come from spatial data analysts wishing to incorporate temporal data in their analysis or vice versa. -->

<!-- This type of data can naturally be considered as a data cube with three axes: spatial identifier, temporal identifier, and variables. In this cubic formation, space only occupies one axis, rather than two, to avoid hypercubes for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the data objects coming to the analysts can vary in forms in the analysis. Figure \ref{fig:illu-input} shows three examples of wild spatio-temporal data: 1) a single table that includes both the spatial and temporal variables; 2) separate tables for spatial and temporal variables; 3) array-based data, i.e. NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid. -->

>>>>

Spatio-temporal data has a spatial component referring to the location of each observation and a temporal component that is recorded at regular or irregular time intervals. It may also include multiple variables measured at each spatial and temporal values. With spatio-temporal data, one can fix the time to explore the spatial features of the data, fix the spatial location/s to explore temporal aspects, or dynamically explore the space and time simultaneously.

In order to computationally explore the spatial, temporal and spatio-temporal faces of such data, the data needs to be stored and represented under a specific data object that allows the user to query, group and dissect all the data faces.

@@

@@

### 2022-06-08 2-the-cubble-pkg.Rmd
proofread

A \code{cubble} object is an S3 class [@advr] built on the \code{tibble} class, specifically to organise spatio-temporal data. The \code{cubble} class uses an attribute "form",  to arrange the spatial or temporal data components tidily. The form attribute can take a value of either "nested" or "long". The nested cubble is a subclass of \code{rowwise_df}. It arranges each spatial site in a row, and uses list columns to store the feature geometry and the temporal information. The long cubble is a subclass of \code{grouped_df}, which expands the temporal information into the long form and stores the spatial information in a "spatial" attribute.

@@

The main functions in the package are \code{as_cubble()}, \code{face_spatial()}, \code{face_temporal()}, and \code{unfold()}. The following sections explain their role, why the new cubble structure is needed and how the package relates to existing packages for spatial and temporal data analysis.

The dataset `climate_flat` is used to illustrate \pkg{cubble} functionality. This is a subset from the ORIGIN and contains the spatial variables, station id, latitude, longitude, elevation, station name, World Meteorology Organisation ID, in additional to daily temporal information,  maximum and minimum temperature values and precipitation records for 2020. The first five rows of `climate_flat` are shown below:

@@

A cubble can be created from various common spatio-temporal data formats, including basic R objects like tibble, tsibble, sf. Section \@ref(covid) describes converting multiple tables into a cubble object and Section \@ref(netcdf) illustrate how to convert a netCDF object.

@@

A cubble object is an extension of a \code{rowwise_df} object where each row forms a group. All the temporal variables are nested in a list column, hence it is also called the nested cubble. The rowwise structure makes it simpler to operate on the list using the \code{mutate()} syntax, which is simpler than the \code{purr::map()} when working with a list column. For example, calculating the number of rainy days can be done by:

@@

The first line in printed cubble now shows the cubble in its long form and the third line has been changed to display the name and type of spatial variables: \code{lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]}. Unlike the nested form, the long cubble is built from a \code{grouped_df} object where all the observations from the same sites form a group.

@@

The \code{dplyr} package has many tools for wrangling tidy data,  many of which are useful in the spatio-temporal analysis. The \code{cubble} package provides \code{methods} that support the use of the following \code{dplyr} operations on both the nested and long forms: `mutate`, `filter`, `summarise`, `select`, `arrange`, `rename`, `left_join`, and the slice family (`slice_head`, `slice_tail`, `slice_sample`, `slice_min`, `slice_max`)

@@

### netcdf {#netcdf}

NetCDF data is another format commonly used for storing spatio-temporal data. It has two main components: *dimension* for defining the spatio-temporal grid (longitude, latitude, and time) and *variable* that populates the defined grid. Attributes  can be associated with dimensions or variables. Because there can be many different styles of representing this information there is a metadata convention to standardise the format of the attributes [@REF(http://cfconventions.org/)]. A few packages in R exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} [@ncdf4], a low-level interface that calls a C-interface: \pkg{RNetCDF} [@rnetcdf], and a tidyverse implementation: \pkg{tidync} [@tidync].

>>>>

A \code{cubble} object is an S3 class [@advr] built on the \code{tibble} class, specifically to organise spatio-temporal data. The \code{cubble} class uses an attribute "form",  to arrange the spatial or temporal data components tidily. The form attribute can take a value of either "nested" or "long". The nested cubble is a subclass of rowwise tibble (\code{rowwise_df}). It arranges each spatial site in a row, and uses list columns to store the feature geometry and the temporal information. The long cubble is a subclass of grouped tibble (\code{grouped_df}), which expands the temporal information into the long form and stores the spatial information in a "spatial" attribute.

@@

The main functions in the package are \code{as_cubble()}, \code{face_spatial()}, \code{face_temporal()}, and \code{unfold()}. The following sections explain their roles, why the new cubble structure is needed and how the package relates to existing packages for spatial and temporal data analysis.

The dataset `climate_flat` is used to illustrate \pkg{cubble} functionality. This is a subset from National Oceanic and Atmospheric Administration (NOAA) Global Historical Climatology Network Daily (GHCND) data. It contains spatial variables, station ID, latitude, longitude, elevation, station name, World Meteorology Organisation ID, in additional to daily temporal information,  maximum and minimum temperature values and precipitation records for year 2020. The first five rows of `climate_flat` are shown below:

@@

\sherry{A cubble can be created from various common spatio-temporal data formats, including basic R objects like tibble, tsibble, sf}. Section \@ref(covid) describes converting multiple tables into a cubble object and Section \@ref(netcdf) illustrates how to convert a netCDF object.

@@

A cubble object is a subclass of the \code{rowwise_df} class where each row forms a group. All the temporal variables are nested in a list column, hence it is also called the nested cubble. The rowwise structure makes it simpler to operate on the list using the \code{mutate()} syntax, which is simpler than the \code{purr::map()} when working with a list column. For example, calculating the number of rainy days can be done by:

@@

The first line in the printed cubble now shows it in the long form and the third line has been changed to display the name and type of spatial variables: \code{lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]}. Unlike the nested form, the long cubble is built from a \code{grouped_df} class where all the observations from the same site form a group.

@@

The \code{dplyr} package has many tools for wrangling tidy data,  many of which are useful in the spatio-temporal analysis. The \code{cubble} package provides \code{methods} that support the use of the following \code{dplyr} operations on both the nested and long forms: `mutate`, `filter`, `summarise`, `select`, `arrange`, `rename`, `left_join`, and the slice family (`slice_*`).

@@

### netCDF {#netcdf}

NetCDF data is another format commonly used for storing spatio-temporal data. It has two main components: *dimension* for defining the spatio-temporal grid (longitude, latitude, and time) and *variable* that populates the defined grid. Attributes  can be associated with dimensions or variables. Because there can be many different styles of representing this information there is a metadata convention [@hassell_data_2017] to standardise the format of the attributes. A few packages in R exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} [@ncdf4], a low-level interface that calls a C-interface: \pkg{RNetCDF} [@rnetcdf], and a tidyverse implementation: \pkg{tidync} [@tidync].

### 2022-06-08 jss/cubble-jss.tex
proofread

Spatio-temporal data has a spatial component referring to the location of each observation and a temporal component that is recorded at regular or irregular time intervals. It may also include multiple variables measured at each spatial and temporal values. With spatio-temporal data, one can fix the time to explore the spatial features of the data, fix the spatial location/s to explore temporal aspects, or dynamically explore the space and time simultaneously.
In order to computationally explore the spatial, temporal and spatio-temporal faces of such data, the data needs to be stored and represented under a specific data object that allows the user to query, group and dissect all the data faces.

@@

A cubble object is an S3 object \citep{advr} built on top of the \code{tibble} class to organise spatio-temporal data. Since spatial and temporal variables are usually recorded in different observational units, the \code{cubble} class uses an attribute, \textbf{form}, to arrange the data into a spatial or temporal tidy way. The \textbf{form} attributes can take a value of either ``nested'' or ``long''. The nested cubble is a subclass of \code{rowwise_df}. It arranges each spatial site in a row and uses list columns to store feature geometry and temporal information. The long cubble, instead, is a subclass of \code{grouped_df}, which expands the temporal information into the long form and stores the spatial information in a unique attribute to the long cubble, \textbf{spatial}.

The main functions in the cubble package are \code{as_cubble()}, \code{face_spatial()}, \code{face_temporal()}, and \code{unfold()}. The following sections explain their role, why the new cubble structure is needed and how the package relates to existing packages for spatial and temporal data analysis.

The dataset \texttt{climate\_flat} is used to illustrate \pkg{cubble} functionality. This is a subset from the ORIGIN and contains the spatial variables, station id, latitude, longitude, elevation, station name, World Meteorology Organisation ID, in additional to daily temporal information, maximum and minimum temperature values and precipitation records for 2020. The first five rows of \texttt{climate\_flat} are shown below:

@@

A cubble can be created from various common spatio-temporal data formats, including basic R objects like tibble, tsibble, sf. Section \ref{covid} describes converting multiple tables into a cubble object and Section \ref{netcdf} illustrate how to convert a netCDF object.

@@

The \code{dplyr} package has introduced many tools for data wrangling tasks and these operations are useful in the spatio-temporal context. \code{cubble} provides methods that support the following \code{dplyr} verbs in both the nested and long form:

\begin{quote}
\texttt{mutate}, \texttt{filter}, \texttt{summarise}, \texttt{select}, \texttt{arrange}, \texttt{rename}, \texttt{left\_join}, and the slice family (\texttt{slice\_head}, \texttt{slice\_tail}, \texttt{slice\_sample}, \texttt{slice\_min}, \texttt{slice\_max})
\end{quote}

@@

\code{tsibble} is a temporal data structure that uses \code{index} and \code{key} to identify the time and different series. \code{cubble} can be seen as following the same vein as \code{tsibble} for spatio-temporal data. This makes it easy to cast a \code{tsibble} into a \code{cubble} as only the \code{coords} argument needs to be supplied:

\begin{CodeChunk}
\begin{CodeInput}
R> # example with a tsibble created from climate_flat
R> raw <- climate_flat |> tsibble::as_tsibble(key = id, index = date)
R> dt <-  raw |> cubble::as_cubble(coords = c(long, lat))
R> dt
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts      
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>  
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_ts>
2 ASN00010311 -31.9  117. 179   york            94623 <tbl_ts>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_ts>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_ts>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_ts>
\end{CodeOutput}
\end{CodeChunk}

When a nested cubble is created, each element in the list-column \code{ts} is of \code{tbl_ts} class and operations available to the tsibble class are still valid within the cubble. For example, the code below calculates two time series features (mean and variance) of the maximum temperature variable:

\begin{CodeChunk}
\begin{CodeInput}
R> # add station-based features in the nested form.
R> dt |> 
+   mutate(
+     fabletools::features(
+       ts, tmax, list(tmax_mean = mean, tmax_var = var)
+       )
+   )
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id          lat  long  elev name  wmo_id ts       tmax_mean tmax_var
  <chr>     <dbl> <dbl> <dbl> <chr>  <dbl> <list>       <dbl>    <dbl>
1 ASN00009~ -31.9  116.  15.4 pert~  94610 <tbl_ts>      25.7    38.6 
2 ASN00010~ -31.9  117. 179   york   94623 <tbl_ts>      26.2    51.1 
3 ASN00010~ -32.9  117. 338   narr~  94627 <tbl_ts>      23.7    45.4 
4 ASN00014~ -12.4  131.  30.4 darw~  94120 <tbl_ts>      33.1     3.02
5 ASN00015~ -17.6  134. 220   elli~  94236 <tbl_ts>      34.6    24.7 
\end{CodeOutput}
\end{CodeChunk}

@@

As a spatial data object, \code{sf} creates a feature geometry list-column (\code{sfc}) in the data frame to provide spatial operations on various geometry types such as (\code{POINT}, \code{LINESTRING}, \code{POLYGON}, \code{MULTIPOLYGON}, etc). When creating a cubble a \code{sf} object can be supplied as the spatial table. Once that is included in a cubble, methods for the \code{sfc} class can be applied in the nested form inside the cubble object. An example of this is shown in \ref{covid}, which also handles the case when the site identifiers in the two tables do not match exactly. A spatial data object with an \code{s2} vector can also be the input for the spatial table in \pkg{cubble}.

\hypertarget{netcdf}{%
\subsubsection{netcdf}\label{netcdf}}

NetCDF data is another format that is commonly used for storing spatio-temporal data. It has two main components: \emph{dimension} for defining the spatio-temporal grid (longitude, latitude, and time) and \emph{variable} that populates the defined grid. Attributes are usually associated with dimensions and variables in the NetCDF format data and a \href{http://cfconventions.org/}{metadata convention for climate and forecast} has been designed to standardise the format of the attributes. A few packages in R exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} \citep{ncdf4}, a low-level interface that calls a C-interface: \pkg{RNetCDF} \citep{rnetcdf}, and a tidyverse implementation: \pkg{tidync} \citep{tidync}.

>>>>

Spatio-temporal data has a spatial component referring to the location of each observation and a temporal component that is recorded at regular or irregular time intervals. It may also include multiple variables measured at each spatial and temporal values. With spatio-temporal data, one can fix the time to explore the spatial features of the data, fix the spatial location/s to explore temporal aspects, or dynamically explore the space and time simultaneously.

In order to computationally explore the spatial, temporal and spatio-temporal faces of such data, the data needs to be stored and represented under a specific data object that allows the user to query, group and dissect all the data faces.

@@

A \code{cubble} object is an S3 class \citep{advr} built on the \code{tibble} class, specifically to organise spatio-temporal data. The \code{cubble} class uses an attribute ``form'', to arrange the spatial or temporal data components tidily. The form attribute can take a value of either ``nested'' or ``long''. The nested cubble is a subclass of rowwise tibble (\code{rowwise_df}). It arranges each spatial site in a row, and uses list columns to store the feature geometry and the temporal information. The long cubble is a subclass of grouped tibble (\code{grouped_df}), which expands the temporal information into the long form and stores the spatial information in a ``spatial'' attribute.

The main functions in the package are \code{as_cubble()}, \code{face_spatial()}, \code{face_temporal()}, and \code{unfold()}. The following sections explain their roles, why the new cubble structure is needed and how the package relates to existing packages for spatial and temporal data analysis.

The dataset \texttt{climate\_flat} is used to illustrate \pkg{cubble} functionality. This is a subset from National Oceanic and Atmospheric Administration (NOAA) Global Historical Climatology Network Daily (GHCND) data. It contains spatial variables, station ID, latitude, longitude, elevation, station name, World Meteorology Organisation ID, in additional to daily temporal information, maximum and minimum temperature values and precipitation records for year 2020. The first five rows of \texttt{climate\_flat} are shown below:

@@

\textcolor{magenta}{A cubble can be created from various common spatio-temporal data formats, including basic R objects like tibble, tsibble, sf}. Section \ref{covid} describes converting multiple tables into a cubble object and Section \ref{netcdf} illustrates how to convert a netCDF object.

@@

The \code{dplyr} package has many tools for wrangling tidy data, many of which are useful in the spatio-temporal analysis. The \code{cubble} package provides \code{methods} that support the use of the following \code{dplyr} operations on both the nested and long forms: \texttt{mutate}, \texttt{filter}, \texttt{summarise}, \texttt{select}, \texttt{arrange}, \texttt{rename}, \texttt{left\_join}, and the slice family (\texttt{slice\_*}).

@@

A \code{tsibble} is a \code{tibble} where the \code{index} and \code{key} components are used to store temporal and strata information, that makes working with temporal data cognitively efficient. A \code{cubble} can use a \code{tsibble} for storing the temporal information, and effectively utilize the specialist time series operations in the \code{tsibble} package. It is easy to cast an existing \code{tsibble} into a \code{cubble} simply by supplying the \code{coords}:

\begin{CodeChunk}
\begin{CodeInput}
R> climate_flat_ts <- climate_flat |> 
+   tsibble::as_tsibble(key = id, index = date)
R> climate_flat_cb <-  climate_flat_ts |> 
+   cubble::as_cubble(coords = c(long, lat))
R> climate_flat_cb
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts      
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>  
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_ts>
2 ASN00010311 -31.9  117. 179   york            94623 <tbl_ts>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_ts>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_ts>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_ts>
\end{CodeOutput}
\end{CodeChunk}

When a nested cubble is created, each element in the list-column \code{ts} is a \code{tsibble} class (labelled \code{tbl_ts}) and operations available to the \code{tsibble} class are still valid on this element. For example, the code below calculates two time series features (mean and variance) of maximum temperature, utilizing the tsibble structure in the cubble:

\begin{CodeChunk}
\begin{CodeInput}
R> # add station-based features in the nested form.
R> climate_flat_cb |> 
+   mutate(
+     fabletools::features(
+       ts, tmax, list(tmax_mean = mean, tmax_var = var)
+       )
+   )
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id          lat  long  elev name  wmo_id ts       tmax_mean tmax_var
  <chr>     <dbl> <dbl> <dbl> <chr>  <dbl> <list>       <dbl>    <dbl>
1 ASN00009~ -31.9  116.  15.4 pert~  94610 <tbl_ts>      25.7    38.6 
2 ASN00010~ -31.9  117. 179   york   94623 <tbl_ts>      26.2    51.1 
3 ASN00010~ -32.9  117. 338   narr~  94627 <tbl_ts>      23.7    45.4 
4 ASN00014~ -12.4  131.  30.4 darw~  94120 <tbl_ts>      33.1     3.02
5 ASN00015~ -17.6  134. 220   elli~  94236 <tbl_ts>      34.6    24.7 
\end{CodeOutput}
\end{CodeChunk}

@@

The \code{sf} is a spatial data object, containing a feature geometry list-column (\code{sfc}) in the form of various geometry types (\code{POINT}, \code{LINESTRING}, \code{POLYGON}, \code{MULTIPOLYGON}). The \code{sf} package provides functions that operate efficiently on this spatial information. A cubble can store the spatial information as an \code{sf} object. Methods for the \code{sfc} class can be applied in the nested form of the cubble object. An illustration is in Section \ref{covid}. A spatial data object with an \code{s2} vector can also be used to store the spatial information in a \code{cubble}.

\hypertarget{netcdf}{%
\subsubsection{netCDF}\label{netcdf}}

NetCDF data is another format commonly used for storing spatio-temporal data. It has two main components: \emph{dimension} for defining the spatio-temporal grid (longitude, latitude, and time) and \emph{variable} that populates the defined grid. Attributes can be associated with dimensions or variables. Because there can be many different styles of representing this information there is a metadata convention \citep{hassell_data_2017} to standardise the format of the attributes. A few packages in R exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} \citep{ncdf4}, a low-level interface that calls a C-interface: \pkg{RNetCDF} \citep{rnetcdf}, and a tidyverse implementation: \pkg{tidync} \citep{tidync}.
